---
title: "Differential Expression Analysis of Indian Dataset"
author: "Solomon"
date: "2023-11-09"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
    theme: "flatly"
editor_options:
  chunk_output_type: console
---

## Indian dataset

We are going to use this dataset as an example to test the working of the 
methods "DESeq2" & "limma"

```{r setup, include=FALSE}
library(SummarizedExperiment)
se <- readRDS("~/tmp/Projects/TestingPhase/indata.rds")
```

## Data processing

```{r setup, include=FALSE}
conditions <- c("Tb_status", "bmi_cat3")
batch <- "batch"
```

## Plotting for each assay {.tabset}

### Counts
```{r setup}
assay_to_analyze <- "counts"

batch <- "batch"
conditions <- c("Tb_status", "bmi_cat2", "BMI")

assay_dat <- assays(se)[[assay_to_analyze]] # Change the data to dat
rownames(assay_dat) <- names(se)
analysis_design <- as.data.frame(colData(se)[c(conditions, batch)])

lim <- limma_method_DE(assay_dat, analysis_design, batch, conditions)
```

## Limma processing

```{r setup, include=FALSE}
# # Define the design matrix
# design_mod <- model.matrix(stats::as.formula(
#     paste(
#         " ~ ",
#         paste(colnames(analysis_design),
#               collapse = "+"
#         )
#     )
# ), data = analysis_design)

library(limma)

# make a new variable, bmi_cat3, that has three levels (low, avg, high bmi)
bmi_cat3 <- rep("avg", ncol(se))
bmi_cat3[colData(se)$BMI < 18.5] <- "low"
bmi_cat3[colData(se)$BMI > 25] <- "high"
colData(se)$bmi_cat3 <- as.factor(bmi_cat3)

# get data matrix
dat_logcpm <- assay(se, "log_counts_cpm") # For limma

# build formula:
mod_formula <- as.formula(paste0(" ~ ", batch, " + ", paste(conditions, collapse = " + ")))
des <- model.matrix(mod_formula, data = colData(se))
colnames(des)[1] <- "Intercept"

# Fit the linear model
fit <- lmFit(dat_logcpm, design = des)

# contrasts for batch
contrasts <- list(levels=des)
contrasts <- append(contrasts, paste(batch, levels(colData(se)[, batch])[-1], sep = "", collapse = " + "))

# contrasts for conditions
for (con in conditions){contrasts <- append(contrasts,paste(con, levels(colData(se)[, con])[-1], sep = "", collapse = " + "))}

# Create the contrast matrix
contrast.matrix <- makeContrasts(contrasts=unlist(contrasts[-1]), levels=des)

# do.call(makeContrasts, myargs)
fit <- contrasts.fit(fit, contrast.matrix)

# Apply empirical Bayes moderation to the standard errors
fit2 <- eBayes(fit)

# Extract top differentially expressed genes
lim <- topTable(fit2, number = Inf)

lim1 <- topTable(fit2, coef = 1)
lim2 <- topTable(fit2, coef = 2, number = Inf)
lim3 <- topTable(fit2, coef = 3)

# Filter genes with adjusted p-value < 0.05 for specified column
topGenes_limma <- subset(lim2, adj.P.Val < 0.05) # Significant gene
topGenes_limma
```

## DESeq2 processing

```{r setup, include=FALSE}
library(DESeq2) # to the top

data <- assays(se)[["counts"]] # For DESeq2
rownames(data) <- names(se)

analysis_design <- as.data.frame(colData(se)[c(conditions, batch)])
colnames(data) <- rownames(analysis_design)
  data[is.na(data)] <- 0

if (any(!data %in% 0:max(data)) || any(data < 0)) {
      stop("Error: The data does contains negative integer counts.")
    }
    dds <- DESeqDataSetFromMatrix(
      countData = data,
      colData = analysis_design,
      design = stats::as.formula(
        paste(
          " ~ ",
          paste(colnames(analysis_design),
            collapse = "+"
          )
        )
      )
    )
    dds <- DESeq(dds)
    dds
```


## TopGenes
### DESeq2

```{r setup, include=FALSE}
perform_analysis_deseq2 <- function(dds, column_name) {

  # Get unique levels of the specified column
  levels <- levels(colData(dds)[[column_name]])
  
  # Check for at least two levels
  if (length(levels) < 2) {
    stop("There must be at least two levels in the specified column for differential expression analysis.")
  }
  
  # Get results for specified column
  res <- results(dds, contrast=c(column_name, levels[1], levels[2]))
  
  # Filter genes with adjusted p-value < 0.05 for specified column
  topGenes <- subset(res, padj < 0.05)
  
  return(topGenes)
}

# Usage
topGenes_deseq2 <- perform_analysis_deseq2(dds = dds, column_name = "Tb_status")
topGenes_deseq2

```

## Venn Diagram representation for the Top genes

Here, I have taken the output of the top genes which have padj value less than 0.05 and plotted a venn diagram representing the genes which might be sharing common space with DESeq2 & limma

```{r setup, include=FALSE}
library(ggvenn)

# Extract the list of genes from the DESeq2 results table
genes_deseq2 <- rownames(topGenes_deseq2)

# Extract the list of genes from the Limma results table
genes_limma <- rownames(topGenes_limma)

# one element for DESeq2 and another for Limma
gene_sets <- list(
  DESeq2 = genes_deseq2,
  Limma = genes_limma
)

# Create and view the Venn diagram
venn_plot <- ggvenn(gene_sets)
print(venn_plot)

```
